Glyph-maps for visualizing climate data and models

### 2011-08-26 paper.tex
included reference to gradual change

Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context. The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map. When measurements are made a different time periods also a common way to display the data is in small multiples, such as in Figure \ref{fig:facetted-map}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America. Color is used to display the deseasonalized temperature measurements, with red being high values and blue being low values. The data comes from \citet{murrell:2010}. The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998. This is the pattern of an El Ni\~no event, a major temperature anomaly. With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years. What the reader is required to do is play the game ``Spot the difference'' from one small image to another. From a cognitive perspective, this is a very hard task to do. It falls into the framework of {\it change blindness} (see e.g. \citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.  Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots.

@@

In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.  For weather maps, icon plots go back as far as \citet{galton:weathermap}. Modern icon plots date back to the semi-graphic display of \citet{anderson:1960} and star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}. An icon is produced by mapping each variable to some feature of the icon. One icon is generated for every case in the data set. Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data. The time measurements are treated as variables in a multivariate data set. Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables. Three different locations are selected, for demontration purposes. The top plot shows the time series of these three locations. Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six eaks. Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected. Location ``3.1'' (blue) has strong seasonality but generally low temperatures. The bottom plot shows these same cases as star glyphs. A separate glyph is created for each location. The 72 time variables form axes organized radially. Each temperature value is plotted relative to the overall range of temperature values. The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon. Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.

>>>>

Climate data is composed of measurements on variables such as temperature, precipitation, and winds that have been collected from a spatial and temporal context. The classic display for spatial data is to use color or grey scale overlaid on geographic location, typically including a map. When measurements are made a different time periods also a common way to display the data is in small multiples, such as in Figure \ref{fig:facetted-map}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a 6 year period of remotely sensed temperature data above Central America. Color is used to display the deseasonalized temperature measurements, with red being high values and blue being low values. The data comes from \citet{murrell:2010}. The most noticeable feature in this display is the strong red in the equatorial region beginning in mid-1997 and tapering out during 1998. This is the pattern of an El Ni\~no event, a major temperature anomaly. With some work other smaller more localized patterns can be perceived, for example, cooler over the land mass in the early year, but warmer in the latter years. What the reader is required to do is play the game ``Spot the difference'' from one small image to another. From a cognitive perspective, this is a very hard task to do. It falls into the framework of {\it change blindness} (see e.g. \citet{healey:2011} and often leads to failure without readers realizing, in particular for small areas of change or subtle changes.  Large structures such as the El Ni\~no event are clear but readers will not be able to effectively glue the images together sufficiently to read off the trend over the years or notice small local deviations between plots. Rendering these small multiples as movie helps with picking up strong trends, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.

@@

In this paper we explore the use of a different type of display, that uses icons on the map, for climate data.  For weather maps, icon plots go back as far as \citet{galton:weathermap}. Modern icon plots date back to the semi-graphic display of \citet{anderson:1960} and star glyphs of  \citep{kleiner:1981}, and were made infamous by \citep{chernoff:1973}. An icon is produced by mapping each variable to some feature of the icon. One icon is generated for every case in the data set. Figure \ref{fig:glyphs} illustrates the idea of icon plots, with emphasis on temporal data. The time measurements are treated as variables in a multivariate data set. Here we have six years of monthly data, giving 72 time values which are considered to be 72 variables. Three different locations are selected, for demonstration purposes. The top plot shows the time series of these three locations. Location ``1.1'' (red) has medium values of temperature, with strong seasonality, the six years corresponding to the six peaks. Location ``2.1'' (green) has generally high temperature values, with weaker seasonality in most years and a lack of low temperatures in the middle of the time period when lows would be expected. Location ``3.1'' (blue) has strong seasonality but generally low temperatures. The bottom plot shows these same cases as star glyphs. A separate glyph is created for each location. The 72 time variables form axes organized radially. Each temperature value is plotted relative to the overall range of temperature values. The seasonal patterns are clearly visible in this type of display, forming six ``petals'' in the icon. Location ``2.1'' has a different pattern than the other two, with only four petals, and location ``3.1'' is a much smaller icon indicating overall lower values.
